# Dates and Times {#sec-datetime}

## {{< fa bullseye >}} Objectives

-   Understand the complexities of working with datetime data
-   Create datetime formatted data from character and numeric encodings
-   Format/print datetime data in the desired format

## Why Dates and Times are hard

I'm going to let Tom Scott deliver this portion of the material for me, as his times and timezones video is excellent and entertaining.

::: youtube-video-container
<iframe src="https://www.youtube.com/embed/-5wpm-gesOY" title="The Problem with Time &amp; Timezones - Computerphile" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen>

</iframe>
:::

There is also an excellent StackOverflow question @freewindWhySubtractingThese2021 and answers [@skeetAnswerWhySubtracting2011,@borgwardtAnswerWhySubtracting2011] demonstrating exactly how times and time zones can get very confusing even in relatively simple circumstances.

Long story short, we will be using libraries in R and python which handle some of these complexities for us, because dates, times, and timezones are **hard** and we *really* don't want to know exactly how hard they are. The libraries I've chosen for this are `datetime` in Python (used by Pandas), and `lubridate` in R.

::: callout-tip
## Try It Out - Getting Set up

::: panel-tabset
### R

```{r, eval = -1}
install.packages("lubridate")
library(lubridate)

# get current date/time
today()
now()
```

[Lubridate cheat sheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) [Lubridate documentation](https://lubridate.tidyverse.org/)

### Python

```{bash, eval = F}
pip install datetime
```

```{python, eval = T}
import datetime

today = datetime.date.today()
today
print(today)

now = datetime.datetime.now()
now
print(now)
```

[pandas datetime documentation](https://pandas.pydata.org/docs/user_guide/timeseries.html)
:::
:::

## Getting Started

Let's differentiate between three types of data which refer to a point in time:

-   a **date**
-   a **time** within a day
-   a **date-time** - a specific time on a specific date

Now, let's think about all of the different ways we can specify dates. The table below has examples along with `strptime` formats that are used in both `R` and `python` for telling the computer which date format is used.

+----------+---------------------+----------+---------------------------------------+-------------------+
|          | Example             | Type     | Notes                                 | `strptime` format |
+=========:+=====================+==========+=======================================+==================:+
| 1        | January 12, 2023    | date     | Common in US/N. America               | %B %d, %Y         |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 2        | 12 January 2023     | date     | Common in Europe                      | %d %B %Y          |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 3        | 01/12/2023          | date     | Common in US                          | %m/%d/%Y          |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 4        | 1/12/23             | date     | Common in US                          | %m/%d/%y          |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 5        | 12/01/2023          | date     | Common in Europe/Asia                 | %d/%m/%Y          |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 6        | 2023-01-12          | date     | ISO 8601 standard\                    | %Y-%m-%d\         |
|          |                     |          | (automatically sorts chronologically) | or %F             |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 7        | 12 2023             | date     | day of year + year                    | %j %Y             |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 8        | 9:23 PM             | time     | 12h time                              | %I:%M %p          |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 9        | 21:23               | time     | 24h time (military time)              | %H:%M\            |
|          |                     |          |                                       | or %R             |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 10       | 21:23:05            | time     | 24h time (with seconds)               | %H:%M:%S\         |
|          |                     |          |                                       | or %T             |
+----------+---------------------+----------+---------------------------------------+-------------------+
| 11       | 2023-01-12T21:23:05 | datetime | ISO 8601 international standard       | %FT%T             |
+----------+---------------------+----------+---------------------------------------+-------------------+

: Different ways to specify dates and times. {#tbl-formats}

Note that rows 4 and 5 of @tbl-formats are ambiguous if you don't know what location your data comes from - the dates could refer to December 1, 2023 or January 12, 2023. This only gets worse if you use 2-digit years.

There are three main ways that you might want to create a date/time @grolemundDatesTimes:

-   From a string
-   From individual date/time components
-   From an existing date/time object

## Creation from Strings

Dates and times are often stored in tabular formats as strings. In some cases, these are read in and automatically formatted as date-times, but in other situations, you have to specify the format yourself.

::: callout-demo
### Demo: Datetimes from Strings

Let's use some data from the US Geological Service with records of earthquakes with magnitude greater than 6 on the Richter scale that occurred between January 1, 2000 and January 1, 2023. You can pull this data yourself using https://earthquake.usgs.gov/earthquakes/map/, but you can also access a CSV of the data [here](https://github.com/srvanderplas/datasets/raw/main/raw/earthquakes2000.csv).

::: panel-tabset
#### R + lubridate

```{r}
library(lubridate)
quake <- read.csv("https://github.com/srvanderplas/datasets/raw/main/raw/earthquakes2000.csv")
str(quake)
```

By default, `read.csv` reads the time information in as a character variable.

```{r, message = F, warning = F}
library(readr)
quake2 <- read_csv("https://github.com/srvanderplas/datasets/raw/main/raw/earthquakes2000.csv")
str(quake2)
```

However, if we use `readr::read_csv`, the data is correctly read in as a `POSIXct` format, which is how R indicates that something is a datetime object.

If we want to directly convert the Time column in `quake` to a datetime, we can use the `lubridate` package, which has helper functions `ymd_hms`, `ymd`, and more. Our data is formatted in ISO 8601 standard format, which means we can easily read it in with `ymd_hms()` .

```{r}
library(lubridate)
library(dplyr)
quake <- quake %>% 
  mutate(dateTime = ymd_hms(Time))
str(quake)
```

We can then test whether `quake$dateTime` is the same as `quake2$Time` :

```{r}
all.equal(quake2$Time, quake$dateTime)
```

So in the case that your data is not automatically read in as a date-time, you can use the helper functions from lubridate (`ymd_hms`, `ymd`, `mdy`, ...) to convert strings to date-time data.

#### Base R

As lovely as the lubridate package is, there are some situations where using the tidyverse may not be desirable or even allowed. It is helpful to know how to solve this problem in base R, even if 99% of the time we can use the much easier-to-remember lubridate package.

In this case, we would use the `as.POSIXct` function, and we probably want to have the reference page up (run `?strptime` in the R console to pull up the help page).

We'll need to get the codes that tell R what format our datetimes use - you can use @tbl-formats, if you like, or read the `as.POSIXct` help page to see all possible format codes.

```{r}
quake <- read.csv("https://github.com/srvanderplas/datasets/raw/main/raw/earthquakes2000.csv")
str(quake)
quake$dateTime2 <- as.POSIXct(quake$Time, "%Y-%m-%dT%H:%M:%S")
all.equal(quake$dateTime, quake$dateTime2)
```

So using `as.POSIXct` we do not get the convenient handling of time zones that we got using `ymd_hms`, but we can set the time zone explicitly if we want to do so.

```{r}
quake$dateTime2 <- as.POSIXct(quake$Time, tz = "UTC", "%Y-%m-%dT%H:%M:%S")
all.equal(quake$dateTime, quake$dateTime2)
```

#### Pandas

In pandas, we can use the `to_datetime` method. If the format is not specified, pandas will try to guess the date-time format; in this case, the guess works, but if not, you can provide a `format = …` argument that works the same way as R.

```{python}
import pandas as pd
quake = pd.read_csv("https://github.com/srvanderplas/datasets/raw/main/raw/earthquakes2000.csv")
quake.dtypes

quake.Time[0:10]

# Convert to datetime
quake['dateTime'] = pd.to_datetime(quake.Time)
quake.dtypes
quake.dateTime[0:10]

# Convert to datetime
quake['dateTime2'] = pd.to_datetime(quake.Time, format = "%Y-%m-%dT%H:%M:%S")
quake.dtypes
quake.dateTime2[0:10]
```
:::
:::

::: callout-tip
### Try it Out - Datetimes from Strings

It's usually important for new parents to keep a log of the new baby's feeds, to ensure that the baby is getting enough liquids and isn't getting dehydrated. 
I used an app to keep track of my daughter's feeds from birth (though here, we'll only work with [the first month of data](https://raw.githubusercontent.com/srvanderplas/datasets/main/raw/feeds_first_month.csv)), and it used a reasonable, if not standard way to store dates and times. 

::: panel-tabset
#### Problem

Take a look at the [first month of feeds](https://raw.githubusercontent.com/srvanderplas/datasets/main/raw/feeds_first_month.csv).
Note that these data are from August 7, 2021 to November 4, 2021 -- roughly baby's first 90 days.

1. Convert Start and End to datetime variables
2. Can you plot the feeds somehow?
3. Can you do arithmetic with datetimes to see if there are any user entry errors?    
This data was created by a *highly* unreliable and error prone couple of individuals -- specifically, sleep-deprived new parents. 

To do this, you may need to figure out how to specify a non-standard date format in R and/or python. The `parse_date_time` function is useful in R, and `pd.to_datetime()` takes a format argument in python.


#### R solution

First, let's read the data in and explore a bit.

```{r}
library(lubridate)
library(readr)
feeds <- read_csv("https://raw.githubusercontent.com/srvanderplas/datasets/main/raw/feeds_initial.csv")
head(feeds)

# Looks like %H:%M:%S %m-%d-%Y format.
```

It looks like the data is stored in a format where the time (`%H:%M:%S`) is first and the date (`%m-%d-%Y`) is second. We can use the `parse_date_time` function in lubridate

```{r}
feeds <- feeds %>%
  mutate(Start = parse_date_time(Start, orders = c("%H:%M:%S %m-%d-%Y")),
         End = parse_date_time(End, orders = c("%H:%M:%S %m-%d-%Y")))
```

Let's then explore how we might plot this data: 

```{r}
#| fig-width: 6
#| fig-height: 1.5
#| fig-caption: "Feeds in the first 90 days of an infant's life."
library(ggplot2)
ggplot(feeds, aes(xmin = Start, xmax = End, fill = Type)) + 
  geom_rect(aes(ymin = 1, ymax = 2)) + # Specify default aes
  scale_fill_manual(values = c("Bottle" = "cornflowerblue", "Breast" = "pink")) + 
  theme_bw() + theme(legend.position = "bottom") + 
  scale_y_continuous(breaks = NULL)
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-caption: "Feeds in the first 90 days of an infant's life, by hour of the day."
library(ggplot2)
feeds %>%
  mutate(day = floor_date(Start, "day"),
         hour_start = Start - day,
         hour_end = End - day) %>%
  mutate(across(starts_with("hour"), ~as.numeric(., units = "hours"))) %>%
  mutate(doy = yday(day)) %>%
ggplot(aes(ymin = day, ymax = day+days(1), xmin = hour_start, xmax = hour_end, fill = Type)) + 
  geom_rect() + # Specify default aes
  scale_fill_manual(values = c("Bottle" = "cornflowerblue", "Breast" = "pink")) + 
  scale_x_continuous("Hour of the day") + 
  theme_bw() + theme(legend.position = "bottom")
```

We can also calculate the duration of each feed and look at the distributions for each type of feed.

```{r}
feeds <- feeds %>%
  mutate(duration = End - Start)

ggplot(feeds, aes(x = duration, fill = Type)) + geom_histogram(color = "black") + 
  scale_fill_manual(values = c("Bottle" = "cornflowerblue", "Breast" = "pink")) + 
  theme_bw() + theme(legend.position = "none") + 
  xlab("Feed duration, in seconds") + facet_wrap(~Type)
```

We can see a few suspiciously long feeds - 9000 seconds is 2.5 hours, which is not unheard of for a baby to breastfeed, but would be an exceptionally long bottle feed (unless a parent fell asleep before hitting "stop" on the feed, which is much more likely). 


#### Python solution


First, let's read the data in and explore a bit.

```{python}
import pandas as pd

feeds = pd.read_csv("https://raw.githubusercontent.com/srvanderplas/datasets/main/raw/feeds_initial.csv")
feeds.head()

# Looks like %H:%M:%S %m-%d-%Y format.
```

It looks like the data is stored in a format where the time (`%H:%M:%S`) is first and the date (`%m-%d-%Y`) is second. We can use the format argument to `pd.to_datetime` to specify this:

```{python}
feeds["Start"] = pd.to_datetime(feeds.Start, format = "%H:%M:%S %m-%d-%Y")
feeds["End"] = pd.to_datetime(feeds.End, format = "%H:%M:%S %m-%d-%Y")
feeds.head()
```

In Python, it is helpful to do a bit of transformation first - this is partly because I'm not as good with Python plotting systems.

```{python}
import datetime as dt
feeds["day"] = feeds.Start.dt.strftime("%Y-%m-%d")
feeds["day"] = pd.to_datetime(feeds.day, format = "%Y-%m-%d")
feeds["day_end"] = feeds.day + dt.timedelta(days = 1)

feeds["time_start"] = feeds.Start - feeds.day
feeds["time_end"] = feeds.End - feeds.day
feeds["duration"] = feeds.time_end - feeds.time_start
```

Note that as of January 2023, RStudio does not correctly display timedelta data types in python.
They show up as NAs in the table, but are printed fine in the console. 
Don't spend hours trying to figure out why it isn't working -- it's bad enough that I did.

```{python}
from plotnine import *

(
  ggplot(feeds, aes(xmin = "Start", xmax = "End", fill = "Type")) + 
  geom_rect(aes(ymin = 1, ymax = 2)) + 
  scale_fill_manual(values = ["cornflowerblue", "pink"]) + 
  theme_bw() + scale_y_continuous(breaks = [])
)

```


```{python}
from plotnine import *

(
  ggplot(feeds, aes(xmin = "time_start", xmax = "time_end", ymin = "day", ymax = "day_end", fill = "Type")) + 
  geom_rect() + 
  scale_fill_manual(values = ["cornflowerblue", "pink"]) + 
  theme_bw()
)

```

:::

:::

## Creation from Components

Sometimes, instead of a single string, you'll have the individual components of the date-time spread across columns. 
The `nycflights13` data is a good example of this.

::: callout-demo
### Demo: Datetimes from Components

::: panel-tabset

#### R + lubridate

In `lubridate`, the `make_date()` and `make_datetime()` functions can be used to create date-times from component pieces.

```{r}
library(nycflights13)

flights %>%
  select(year, month, day, hour, minute) %>% 
  head()

flights <- flights %>%
  mutate(date = make_date(year, month, day),
         datetime = make_datetime(year, month, day, hour, minute))

flights %>% select(date, datetime, year, month, day, hour, minute)
```

#### Base R

In base R, we can use the `ISOdate` function to create date times.

```{r}
flights$datetime_base = with(flights, ISOdatetime(year, month, day, hour, minute, sec= 0, tz="UTC"))
all.equal(flights$datetime, flights$datetime_base)
```

#### Python

In pandas, we can pass multiple columns to `pd.to_datetime()` and as long as they are named reasonably, pandas will handle the conversion.
If we want to have the date but not the time for some reason, we just pass fewer columns to pandas.

```{python}
from nycflights13 import flights

flights[["year", "month", "day", "hour", "minute"]]

flights["date"] = pd.to_datetime(flights[["year", "month", "day"]])
flights["datetime"] = pd.to_datetime(flights[["year", "month", "day", "hour", "minute"]])


flights[["date", "datetime", "year", "month", "day", "hour", "minute"]]
```

:::
:::


## Creation from other objects

Another way to create date-time objects is to modify other objects. 